Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-03-12 Manuscript/Figures/figure_legends.Rmd
current

**Figure 1: Model Construction.** An illustration of the proposed model construction given cell line-specific Hi-C matrices obtained from [@rao20143d]. The contact matrix above is a visualization of the interconnectivity of chromosome 14 (GM12878) on the intervals from 20 mb to 100 mb [@durand2016juicebox]. The coordinates of subsequent topologically association domains were identified using the arrowhead algorithm from Juicer. The linear genome, represented by the diagonal of the contact matrix, was then binned according to prespecified interaction frequency resolutions (10 kb, 25 kb, 50 kb, and 100 kb). Each TAD coordinate was then flanked by a unit-specific amount (in basepairs) respective to the resolution of the corresponding contact matrix. The response vector Y, used for classification, was determined by whether a genomic bin overlapped with a flanked boundary region (Y=1) or not (Y=0). The spatial relationship between positional coordinates of various functional genomic annotations, obtained from ENCODE, and the genomic bins were then used to define the feature space of the models.

>>>>

**Figure 1: Model Construction.** An illustration of the proposed model construction given cell line-specific Hi-C matrices. The contact matrix above is a visualization of the interconnectivity of chromosome 14 (GM12878 cell line, [???] resolution) on the intervals from 20 mb to 100 mb. The boundary coordinates of topologically association domains (TADs) were identified using the Arrowhead algorithm. To account for the uncertainty in TAD boundary detection, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$, used for classification, was determined by whether a genomic bin overlapped with a flanked boundary region (Y=1) or not (Y=0). The spatial relationship between positional coordinates of various functional genomic annotations, obtained from ENCODE, and the genomic bins were then used to define the feature space of the models.

### 2019-03-12 Manuscript/manuscript.Rmd
current

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used <!--to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->.The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

We evaluated and compared multiple random forest classification algorithms built on domain data extracted from Hi-C datasets for 2 cell lines C = {GM12878, K562}, at 4 different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}. For each classifier, we compared 3 predictor types P = {OC, OP, Distance} and 5 class balancing techniques B = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {c, r, p, b}, where $c \in C, \quad R \in R, \quad p \in P,$ and $b \in$ B. Letting |X| denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling X (=$x \times 4 \times 3 \times 5$) separate models. For each model, we aimed to assess both model performance and identify the genomic annotations that were most important in correctly labeling the genomic bins that overlapped with flanked TAD boundaries.

@@

We assessed the collinearity among the genomic annotations used for the feature space of subsequent models by isolating the genomic bins that overlapped with flanked TAD boundaries. We then vectorized the number of instances when a region defined by a genomic annotation overlapped with the isolated bins. It was found that the correlations among these vectors were very high for many genomic annotations. The correlations were visualized using correlograms for each data resolution and cell line (Supplementary Figure 2). The high correlations imply that many genomic annotations tended to colocalize near eachother at the boarders of domains. This was present across both resolution and cell line. Specifically, we found that the CTCF, Rad21, Smc3ab, and Znf143 transcription factor binding sites were highly colocalized near TAD boundaries. Likewise, many histone modifications also tended to appear near TAD boundaries in conjunction with one another including H3K4me1, H3K4me2, H3K4me3, H3K9ac, and H3K27ac. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated with on another, yet inversely correlated with many transcription factor binding sites.

The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed after implementing each of the resampling techniques on all data resolutions within each cell line. Additionally, a unique predictor-type was included separately as the set of features. It was found that the elastic-net regularized the coefficients of far more distance-type predictors to 0 compared to datasets with overlap count and percent predictor types, regardless of cell line or data resolution (Supplementary Figure 3). Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.

>>>>

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used <!--to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->.The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were recorded.

@@

We evaluated and compared multiple random forest classification algorithms built on domain data extracted from Hi-C datasets for 2 cell lines $C = {GM12878, K562}$, at four different resolutions $R = {10 kb, 25 kb, 50 kb, 100 kb}$. For each classifier, we compared three predictor types $P = {OC, OP, Distance}$ and five class balancing techniques $B = {None, RUS, ROS, SMOTE, ROSE}$. A system produced from the ensemble framework was defined as: $ = {c, r, p, b}$, where $c \in C, \quad R \in R, \quad p \in P,$ and $b \in B$. Letting $|X|$ denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $X (= x \times 4 \times 3 \times 5)$ separate models. For each model, we aimed to assess both model performance and identify the genomic annotations that were most important in correctly labeling the genomic bins that overlapped with flanked TAD boundaries.

@@

We assessed the collinearity among the genomic annotations used for the feature space of subsequent models by isolating the genomic bins that overlapped with flanked TAD boundaries. We then vectorized the number of instances when a region defined by a genomic annotation overlapped with the isolated bins. It was found that the correlations among these vectors were very high for many genomic annotations. The correlations were visualized using correlograms for each data resolution and cell line (Supplementary Figure 2). The high correlations imply that many genomic annotations tended to colocalize near each other at the boarders of domains. This was present across both resolution and cell line. Specifically, we found that the CTCF, Rad21, Smc3ab, and Znf143 transcription factor binding sites were highly colocalized near TAD boundaries. [??? We need to quantify and show this, and the following colocalizations] Likewise, many histone modifications also tended to appear near TAD boundaries in conjunction with one another including H3K4me1, H3K4me2, H3K4me3, H3K9ac, and H3K27ac. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites.

The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed after implementing each of the resampling techniques on all data resolutions within each cell line. Additionally, a unique predictor-type was included separately as the set of features. It was found that the elastic-net regularized the coefficients of far more [??? quantify] distance-type predictors to 0 compared to datasets with overlap count and percent predictor types, regardless of cell line or data resolution (Supplementary Figure 3). Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.
